Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-22 paper.tex

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata~\citep{danecek2011variant}. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
Our Future Health~\citep{ofhmain_2025},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
Large genetic variation datasets are also being generated for other organisms
and a variety of purposes including 
agriculture~\citep{ros2020accuracy,wang2023rice}, 
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats 
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN formats are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits 
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

@@

\subsection{Case study: Genomics England WGS data}
In this section we demonstrate the utility of VCF Zarr on a large human dataset
and the scalability of the \texttt{vcf2zarr} conversion utility.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
GWAS~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

@@

\subsection{Case study: Our Future Health genotype data}
Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
It is designed to help people live healthier lives for longer through the 
discovery and testing of more effective approaches to prevention, earlier 
detection and treatment of diseases.
Our Future Health is inviting millions of people, from all backgrounds and 
from right across the UK to take part. Volunteers are providing information 
about their health and lifestyles to create an incredibly detailed picture 
that represents the whole of the UK.

Release 9~\citep{ofhdatarelease9}
was made available in December 2024. The genetic component includes genotype data 
from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
10,221 variants.

@@

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-top-fields-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
% /call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
% /call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
% /call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
% /call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
% /call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
% /call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
% /call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
\bottomrule
\end{tabular}
\end{table}

@@

This study makes use of de-identified data held by Our Future Health. 
We would like to acknowledge all the research participants who have donated 
their data to the Our Future Health research programme~\cite{ofhpubpolicy}.

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata~\citep{danecek2011variant}. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
Large genetic variation datasets are also being generated for other organisms
and a variety of purposes including 
agriculture~\citep{ros2020accuracy,wang2023rice}, 
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats 
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN formats are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits 
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

@@

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of VCF Zarr on a large human dataset
and the scalability of the \texttt{vcf2zarr} conversion utility.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
GWAS~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

@@

\subsection{Case study: Our Future Health genotype release}
Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
It is designed to help people live healthier lives for longer through the 
discovery and testing of more effective approaches to prevention, earlier 
detection and treatment of diseases.
Our Future Health is inviting millions of people, from all backgrounds and 
from right across the UK to take part. Volunteers are providing information 
about their health and lifestyles to create an incredibly detailed picture 
that represents the whole of the UK.

% this section is probably better placed in acknowledgements?
This study makes use of de-identified data held by Our Future Health. 
We would like to acknowledge all the research participants who have donated 
their data to the Our Future Health research programme 
\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}

Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
was made available in December 2024. The genetic component includes genotype data 
from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
10,221 variants.

@@

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-top-fields-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
\bottomrule
\end{tabular}
\end{table}

@@
